Neural Connectome of the Ctenophore Statocyst

### 2025-07-10 manuscript/Manuscript.qmd
Ms: edit results, discussion, add line breaks for easier overview of changes

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding. From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. These were imaged using a Zeiss Gemini 500 SEM at a resolution of 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016]. We reconstructed 972 cells with intact nuclei and somata. Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons. Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume. We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, epithelial floor cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figure supplement 1). These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.

@@

To investigate whether balancer cilia are coordinated by neural inputs, we performed high-speed video microscopy to analyze their activity patterns across body axes. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. In all cases, only larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation) were included for analysis. This ensured that all balancer cilia experienced comparable gravitational input from the statolith, allowing for an unbiased comparison across quadrants.
We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

Notably, bridge neurons associated with each quadrant region (Q1Q2 or Q3Q4) extend their processes across the midline, forming connections that span between opposite quadrant domains when viewed along the tentacular plane. This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination. By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating information flow across orthogonal planes, thus supporting coordinated activity throughout the aboral organ.

@@

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei. The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. A recent study also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@Ferraioli2025]. However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN. Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion. The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.

ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells, bridge cells or other sensory cells. This suggests that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].

>>>>

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. 
To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding. 
From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. 
These were imaged using a Zeiss Gemini 500 SEM at a pixel size (xy) of 2.0 nm. 
We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. 
The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. 
We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016]. 
We reconstructed 972 cells with intact nuclei and somata. 
Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons. 
Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume.
We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, epithelial floor cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figure supplement 1). 
None of these other cell types synapsed on the ANN.
These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.

@@

To investigate whether balancer cilia are coordinated by neural inputs, we did high-speed video microscopy recordings to analyze their activity patterns across body axes. 
Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. 
Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. 
To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. 
In all cases, only larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation) were included for analysis. 
This ensured that all balancer cilia experienced comparable gravitational input from the statolith, allowing for an unbiased comparison across quadrants.
We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

Notably, bridge neurons associated with each quadrant region (Q1Q2 or Q3Q4) extend their processes across the midline, forming connections that span between opposite quadrant domains when viewed along the tentacular plane. 
This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination. 
By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating the spread of activity across orthogonal planes, thus supporting coordinated beating throughout the aboral organ.

@@

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. 
Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei. 
The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. 
A parallel vEM study of a one-day-old cydippid also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@ferraioli2025]. 
However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN in the earlier stage. 
Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion. 
The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.

ANNs form no synaptic output to classical effector organs such as muscles or comb plates. 
All observed output synapses were directed to balancer ciliary cells, bridge cells or other sensory cells. 
This suggests that ANNs are unlikely to serve as motor drivers (premotor neurons or motoneurons) in the conventional sense. 
Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].
